\documentclass [12pt,letterpaper]{report}

% Standard packages
\usepackage{amsmath}		% Extra math definitions
\usepackage{graphics}		% PostScript figures
\usepackage{setspace}		% 1.5 spacing
\usepackage{longtable}          % Tables spanning pages
\usepackage{verbatim}

% Vijay packages
\usepackage{url}
\usepackage{graphicx}
\usepackage[table]{xcolor}
\usepackage[lofdepth,lotdepth]{subfig}

% Custom packages
\usepackage[first]{datestamp}	% Datestamp on first page of each chapter
\usepackage[fancyhdr]{McECEThesis}	% Thesis style
\usepackage{McGillLogo}		% McGill University crest

% $Id: ThesisEx.tex,v 1.1 2005/06/09 12:48:46 kabal Exp $

\usepackage{color}
\def\headrulehook{\color{red}}		% Color the header rule

%===== page layout
% Define the side margins for a right-side page
\insidemargin = 1.3in
\outsidemargin = 0.8in

% Above margin is space above the header
% Below margin is space below footer
\abovemargin = 1.1in
\belowmargin = 0.75in

%========= Document start

\begin {document}

%===== Title page

\title{A Tool for Configuring Mappings for Musical Systems using Wireless Sensor Networks}
\author{Vijay Rudraraju}
\date{\Month\ \number\year}
\organization{%
  \\[0.05in]
  \McGillCrest {!}{1in}\\	% McGill University crest
  \\[0.05in]
  Music Technology Area\\
  Schulich School of Music\\
  McGill University\\
  Montreal, Canada}
\note{%
  {\color{red} \hrule height 0.4ex}
  \vskip 2ex
  A thesis submitted to McGill University in partial fulfillment of the
  requirements for the degree of Master of Arts.
  \vskip 2ex
  \copyright\ \the\year\ Vijay Rudraraju
}

\maketitle

%===== Justification, spacing for the main text
\raggedbottom
\onehalfspacing
\pagenumbering{roman}

%===== Abstract, Sommaire & Acknowledgments
\section*{\centering Abstract}

Digital musical instruments, which are defined here as interactive musical systems containing a control mechanism and a sound generation mechanism, are powerful tools for analyzing performance practice and for transforming and reimagining the bounds of musical performance. However, the transitory nature of digital technology and the complexity of maintaining and configuring a digital musical instrument involving tens, if not hundreds, of interconnected, discrete components presents a unique problem. 

Even the most mechanically complex acoustic musical instruments, like a piano, are robust enough to withstand the daily grind without expert intervention by someone with intimate knowledge of the material and mechanical construction of the instrument. Furthermore, they are standardized enough that repairs can be conducted by any number of trained professionals. By contrast, digital musical instruments are often configured differently for each performance (this configurability being one of the virtues of a digital musical instrument), incorporate any number of non-standard pieces of hardware and software, and often can only be reliably configured by their creator.

This problem is exacerbated as the number of sensors that make up the control mechanism in an instrument increases and the interaction of the control mechanism with the sound generation mechanism grows more complex. This relationship between the control mechanism and the sound generation mechanism is referred to here as the "mapping" of the instrument. The mapping for an instrument represents the aspect of an instrument that is usually most configurable because it is defined by software (as opposed to hardware) and also most crucial to the character of the instrument. In the case of a digital musical instrument, being able to easily configure the musical instrument becomes a point of artistic freedom in addition to a point of maintainability.

This thesis builds upon work encompassed in two projects at the Input Devices and Musical Interaction Lab, the Digital Orchestra Project and Libmapper, to tackle the problem of building an interface/system for configuring a complex musical system without expert programming skills. The intent is to present a targeted survey of user interface design and data visualization design research through the years to inform the design of a graphical user interface for performing this configuration task.

\section*{\centering R\'{e}sum\'{e}}

Les instruments de musique num\'{e}riques, que l'on d\'{e}finit ici comme des syst\`{e}mes musicaux interactifs contenant un m\'{e}canisme de contr\^{o}le ainsi qu'un m\'{e}canisme de production sonore, constituent de puissants outils pour l'analyse des m\'{e}thodes de performance ainsi que pour la transformation et la red\'{e}finition de la la performance musicale. Cependant, la nature \'{e}ph\'{e}m\`{e}re des technologie digitales ainsi que la complexit\'{e} du maintien de la configuration d'un instrument de musique num\'{e}rique, comprenant des dizaines voire centaines de composantes discr\`{e}tes interconnect\'{e}es, pr\'{e}sente un r\'{e}el probl\`{e}me. 

M\^{e}me les syst\`{e}mes musicaux acoustiques le plus complexes, comme le piano, sont assez robustes pour r\'{e}sister \`{a} l'usure quotidienne sans l'intervention d'une main experte ayant une connaissance aiguis\'{e}e des mat\'{e}riaux et de la m\'{e}canique de l'instrument. Par ailleurs, ces instruments sont assez standards pour que tout luthier qualifi\'{e} soit en mesure d'intervenir sur l'instrument en cas de besoin. A l'inverse, les instruments de musique num\'{e}riques pr\'{e}sentent souvent des configurations propres a chaque performance (cette configurabilit\'{e} \'{e}tant une des vertus des instruments digitaux), incluent divers type de syst\`{e}me hardware et software, et ne peuvent souvent \^{e}tre configur\'{e}s de mani\`{e}re fiable que par leur cr\'{e}ateur.

Le probl\`{e}me grandit lorsque le nombre de capteurs utilis\'{e}s pour la partie contr\^{o}le d'un instrument augmente et que l'interaction entre les sous-blocs contr\^{o}le et production sonore se complexifient. Cette relation entre le m\'{e}canisme de contr\^{o}le et de production sonore est appel\'{e}e "mapping" de l'instrument. Le "mapping" d'un instrument se rapporte aux aspects les plus configurables d\'{e}finis par le logiciel (par opposition au \'{e}l\'{e}ments hardware), et mais aussi faisant partie les caract\'{e}ristiques les plus importantes de l'instrument. Dans le cas d'un instrument de musique num\'{e}rique, la capacit\'{e} a agir facilement sur la configuration devient alors un enjeux de libert\'{e} d'expression artistique, en plus d'une condition de p\'{e}rennisation.

Ce travail de th\`{e}se se construit sur la base de deux projets men\'{e}s au sein du laboratoire IDMIL (Input Devices and Musical Interaction Lab), le "Digital Orchestra" et "Libmapper", et traite du probl\`{e}me de construction d'un syst\`{e}me/interface pour la configuration de syst\`{e}me musicaux complexes sans connaissances expertes en programmation. L'objectif est de pr\'{e}senter un sondage cibl\'{e} sur la recherche en conception d'interface utilisateur et visualisation de donn\'{e}es, dans le but d'apporter des \'{e}l\'{e}ments nouveaux a la conception d'interfaces destin\'{e}es aux taches de configuration d'instruments digitaux.

\newpage

\section*{\centering Acknowledgments}

Many thanks to my supervisor, Professor Marcelo M. Wanderley, for his abundance of available time for his students and helping me integrate my personal work with the work of others at the Input Devices and Music Interaction Laboratory (IDMIL).

I am grateful to have been in the presence of all the students and professors of Music Technology at McGill University. Never have I been part of a group of people who were so eager to offer their time to others who needed help, regardless of whether they had something to gain from it. Special thanks to Joseph Malloch and Stephen Sinclair, without whose help and hours of work on the Mapper Tools project I would never have been able to even contemplate this research project.

I would never have survived the Montreal winters if it was not for Vincent Freour, a boundless source of energy who pulled me out into the sunlight even when I was determined to remain in my cave.

Finally, infinite gratitude to my parents Pandu and Padma who have always and will continue to support me, no matter how seemingly useless and lost I may appear through my many journeys.

%========== Tables of contents, figures, tables
\tableofcontents
\listoffigures
\listoftables

\newpage
\chapter*{List of Acronyms}\markright{List of Terms}

\begin{longtable}{ll}
  HCI		&	Human-Computer Interaction\\
  UID		&	User Interface Design\\
  DMI		&	Digital Musical Instrument\\
  EMI		&	Electronic Musical Instrument\\
  OSC	&	Open Sound Control\\
  IDMIL	&	Input Devices and Music Interaction Laboratory\\
\end{longtable}

\cleardoublepage
\pagenumbering{arabic}

%========== Chapters
\typeout{}
\include{introduction}

%==========
\typeout{}
\include{mapping}

%==========
\typeout{}
\include{user_interface_and_data_visualization_design}

%==========
\typeout{}
\include{vizmapper}

%==========
\typeout{}
\include{conclusions}

%========== Appendices
\appendix
\include{resources}
\include{definitions}

%==========
\begin{comment}
\typeout{}
\input{A-A}
\end{comment}

%========== Bibliography
\typeout{}
\begin{singlespace}
  \bibliography{vijay_thesis}
  \bibliographystyle{ieeetr}
\end{singlespace}

\end{document}
